---
title: "NCRMP Benthics statistical package in R"
author: "Sarah Groves, Bethany Williams, Shay Viehman"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
      toc: true
      toc_depth: 2

vignette: >
  %\VignetteIndexEntry{NCRMP Benthics statistical package in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)

library(magrittr)
library(tidyverse)

```

# Description 

The ncrmp.benthic.analysis package is designed to use benthic coral reef data collected from the Atlantic regions of NOAA's National Coral Reef Monitoring Program (NCRMP) to complete the following:

1. Compute summary statistics (e.g. coral density, percent cover, species richness, old and recent mortality, disease prevalence, *Diadema* density) from the site to the regional level as well as presence/absence of ESA species and bleaching abundance at the strata and site levels, respectively.

2. House all current NCRMP benthic data ranging from analysis ready data to regional domain estimates in a standardized format. This includes all metrics listed in item #1 calculated at the site, strata and regional level (as applicable) as well as the most recent sampling grids and NTOT files. There are over 250 dataframes stored in this package.

Note: 

* This package assumes some familiarity with NCRMP sampling methods and analysis ready (AR) data formats. NCEI and CoRIS archived AR data, data dictionaries and sampling protocols can be found on the NCRMP project page, https://coastalscience.noaa.gov/project/national-coral-reef-monitoring-program-biological-socioeconomic/ under "Products, Datasets and Reports"


* This package currently uses NOAA's NCRMP Atlantic benthic data only. See https://github.com/jeremiaheb/rvc#reef-visual-census-statistical-package-in-r for Atlantic fish R package.

* Where indicated in file names by NCRMP_FRRP or NCRMP_DRM, NCRMP and Disturbrance Response Monitoring (DRM; formerly known as Florida Reef Resilience Program or FRRP) data from 2014 to present have been combined during co-sampling (even) years. DRM only data are used for non-NCRMP (odd) years. We are in the process of changing the naming convention from FRRP to DRM but at present DRM and FRRP are used interchangeably. Please be patient while we complete this transition. 

* SEFCRI 2014 coral demographics, benthic cover and invertebrates/ESA corals data, FLK 2014 benthic cover and invertebrates/ESA corals data, and Tortugas 2018 coral demographic data are 2 stage (2 transects per site). Site level metrics for these regions and years are the mean of transect 1 and transect 2. Strata and regional means are then calculated from these site level data. 


For questions, please contact the repository owner, Shay Viehman, shay.viehman@noaa.gov 


### NCRMP Regions and sampling years

* Florida:  
    * Southeast Florida (SEFCRI; 2014, 2016, 2018, 2020, 2022)
    * Florida Keys (FLK; 2014, 2016, 2018, 2020*, 2022)
    * Dry Tortugas (Tortugas; 2014, 2016, 2018, 2020, 2022) 

* Caribbean:
    * St. Thomas & St. John (STTSTJ; 2013, 2015, 2017, 2019, 2021)
    * St. Croix (STX; 2015, 2017, 2019, 2021)
    * Puerto Rico (PRICO; 2014, 2016, 2019, 2021)

* Gulf of Mexico:
    * Flower Garden Banks National Marine Sanctuary (GOM; 2013, 2015, 2018, 2022)
    
Upcoming data additions: Caribbean 2023 (in 2024)


## Installation 

1. Download the zipped R package using the green "Clone or download" button. Unzip the file. 

2. In R studio, open the ncrmp.benthics.analysis R project in the folder of the same name and use the following code to install the pacakge. You must be in the in the ncrmp.benthics.analysis R project. If you aleady have devtools installed you can skip the first line. 

```
install.packages('devtools')

devtools::install()
```
3. Exit the ncrmp.benthics.analysis R project and check your R library to confirm the package installed. At this point you can either delete the files from the zipped folder or retain them if you are interested in the analysis code. 

4. From your current project, use the following code to load and use the package. 

```
library(ncrmp.benthics.analysis) 
```
5. To update the package (when new data is added), repeat steps 1-4. If you are having trouble, try deleting the old version first before intalling the new one. 

Note: Requires R version 3.4.1 or higher.

Currently, you cannot load the package directly from Git Hub using devtools::install_github('shgroves/NCRMP_benthics'). Future updates will inlcude this feature. 

______________________________________________________________________________________________________


# To use this package as a data repository 

To use this package as a data repository requires very little experience with R! You will need to have R version 3.4.1 or higher installed on your computer as well as R Studio desktop. Please see https://cran.r-project.org/doc/manuals/r-release/R-admin.html and  https://www.rstudio.com/products/rstudio/download/ for more information. 

Because all datasets are stored in the R package, all that is required to access these dataframes is to load the package, and type the name of the dataset. You can then begin using it in your code or you can assign it to an object if you want to store it in your Global Environment.



__Examples:__

__Export a dataframe__
```{r}
# require(ncrmp.benthics.analysis)

# write.csv(NCRMP_DRM_FLK_2014_18_density_site, "Your file path\\NCRMP_DRM_FLK_14_16_density_site.csv", row.names = F)

```

__Use a dataframe in a ggplot__
```{r}
require(ncrmp.benthics.analysis)
require(ggplot2)

# Load weighted strata means
Hardcoral <- NCRMP_STTSTJ_2013_21_cover_region %>%
  dplyr::ungroup() %>%
  # filter to just hardcoral
  dplyr::filter(cover_group == "HARD CORALS") %>%
  # Change Year to factor so R sees it as a grouping variable and not a number
  dplyr::mutate(YEAR = as.factor(YEAR))

# Create your plot
ggplot(Hardcoral, aes(x = YEAR, y = avCvr, group = YEAR)) +
  geom_bar(aes(fill = YEAR), stat="identity", colour="black",
           position=position_dodge(),
           size=.3) +
# Add error bars
  geom_errorbar(aes(ymax = avCvr+SE, ymin = avCvr-SE), width = .5, position=position_dodge(1)) +
# Set the y axis scale
  scale_y_continuous(limits = c(0,25), expand = c(0, 0)) +
# Add axis labels
  xlab("Year") +
  ylab("Cover (%)") +
  labs(title = "Hard coral cover") + 
# Remove legend 
  theme(legend.position="none")

```


__Add dataframe to the Global Environment__
```{r}
# require(ncrmp.benthics.analysis)
#
# #The 'data' function will load a dataframe to the Global Environment keeping the existing name, 
# 
# data("NCRMP_DRM_FLK_2014_22_density_site")
#
# or you can give it a new name 
#
# dat <- NCRMP_DRM_FLK_2014_22_density_site

```


## Datasets 

Datasets contained within the package at the strata and regional level are created using the default function parameters: 

  * For Florida regions:
      * project = "NCRMP" (NCRMP only) or "NCRMP_DRM" (NCRMP and DRM combined)
      * species filter = FALSE (filter not applied)
     
      
      
  * For all Carib regions: 
      * project = "NCRMP" 
      * species filter = FALSE 
      
      
      
  * For GOM:
      * project = "NCRMP" 
      * species filter = FALSE 
      
  
### Analysis Ready (AR) data 

Analysis ready NCRMP datasets for all regions and years are contained within this package in the most recent NCRMP Atlantic archive format. AR datasets can be identified because they begin with the region code followed by the year and then the data type; benthic_cover, coral_demographics or inverts_ESAcorals. Two stage (2 transect) data are identified in the file name except for the Tortugas 2018 data. Tortugas 2018 coral demographics surveys were a joint mission between NCRMP and DRM where two transects were conducted at each site. The NCRMP and DRM data are combined and indistinguishable. All other datasets are single transect. 

* Data format: 
    * REGION_YEAR_datatype
  
* Examples: 
    * FLK_2016_coral_demographics,
    * SEFCRI_2014_2stage_benthic_cover                    
    * FGBNMS_2013_inverts_ESAcorals


Additionally, there are two datasets that combine all regions and years into a single dataframe. 
* NCRMP_benthic_cover_All_regions_years
* NCRMP_demo_All_regions_years
  
   
In addition, the following AR Florida DRM and Sanctuary Coral Reef Ecosystem Assessment Monitoring (SCREAM) datasets in NCRMP format are stored in the package. Combined DRM and SCREAM data are from 1999 to 2013 (not all regions are sampled each year) and individual DRM regions are 2014 to the most recent year of sampling.

* preNCRMP_SCREAM_DRM_coral_demographics  
* DRM_SEFCRI_2014_2022_2stage_coral_demographics - includes data from 2014-2022 
* DRM_FLK_2014_2022_2stage_coral_demographics - includes data from 2014-2022 
* DRM_Tort_2014_2022_2stage_coral_demographics - includes data from 2014-2022 

### Summary dataframes 

Summary dataframes are provided for site, strata and weighted regional means. Every function listed below can produce several dataframes. Although there are too many to list individually, all provided dataframes follow a standard naming convention. Summary dataframes can be identified because they begin with the project or projects (NCRMP or NCRMP_DRM), followed by the region, year(s) and data type and level (site, strata, region). See associated functions used to create these dataframes (below) for more information. To view all of the data contained within this package you can 1. download the package from GitHub and view the 'data' folder or 2. use the following code to view all of the datasets in R.

```{r}

# data(package='ncrmp.benthics.analysis')

```


* Data format: 
    * PROJECT_REGION_YEAR_datatype_level
  
* Examples: 
    * NCRMP_DRM_FLK_2014_22_dis_prev_site, 
    * NCRMP_FGBNMS_2013_22_density_strata,                     
    * NCRMP_FLK_2014_22_rec_mort_region,
    * NCRMP_PRICO_2014_21_richness_strata,
    * NCRMP_STX_2015_21_cover_region
    
In addition, there are three dataframes that contain all regions and years

* NCRMP_AllRegions_Years_ESA_PresAbs_Site
* NCRMP_AllRegions_Years_ESA_PresAbs_Strat
* NCRMP_bleaching_abundance_Allyears_Allregions
            
### Shapefiles and NTOT files 

The most recent sampling grid shapefiles used for site allocation and all NTOT (grid summary) files are also contained in the package. Grids can be used in R or exported as shapefiles. The NTOT files are required to properly apply the weighting scheme to summary data and often differ between years. The Florida Keys National Marine Sanctuary (FKNMS) and the Flower Garden Banks National Marine Sanctuary (FGBNMS) boundary polygons are also contained in the package. More boundary polygons for other regions will be added in future updates.

* Sampling grids:
    * SEFCRI_2018_sample_frame
    * FLK_2020_coral_demographics
    * Tort_2020_sample_frame
    * STTSTJ_2021_sample_frame_WGS1984
    * STX_2021_sample_frame_WGS1984
    * PRICO_2021_sample_frame_WGS1984
    * FGBNMS_2018_sample_frame_WGS1984

* NTOT files:
    * USVI_2021_NTOT (STTSTJ & STX)
    * FLK: FLK_2014_NTOT, FLK_2016_NTOT, FLK_2018_NTOT, FLK_2020_NTOT, FLK_2022_NTOT
    * SEFCRI: SEFL_2014_NTOT, SEFL_2016_NTOT, SEFL_2018_NTOT, SEFL_2020_NTOT, SEFL_2020_NTOT
    * Tortugas: Tort_2014_NTOT, Tort_2016_NTOT, Tort_2018_NTOT, Tort_2020_NTOT, Tort_2022_NTOT
    * PRICO_2021_NTOT
    * FGBNMS_2022_NTOT

* Boundaries
    * FKNMS_boundary 
    * FGBNMS_boundary


### Additional files

* ncrmp_frrp_sppcodes: species code, species names and benthic categories for NCRMP and other Florida monitoring programs. 

# To use the functions in this package

If you would like to run the AR data through the functions, apply preset species filters to certain functions or just know more about how NCRMP metrics are calculated, please feel free to use these functions or check out the code. Comments or suggestions on how to improve the package are welcome. 

# Functions for specific regions

### These functions create outputs that contain a single region and either single or multiple sampling years. They all use a stratified random sampling weighting scheme to calculate regional (sampling domain) level metrics. 


## Benthic cover

### 1a. NCRMP_calculate_cover

__Calculates percent cover from species to regional level for individual years__


* parameters:
    * region: "SEFCRI", "FLK", "Tortugas", "STTSTJ", "STX", "PRICO", "GOM" 

    * project: "NCRMP", or "MIR" Note - MIR option is in development. 

* outputs (dataframes):
    * cover_group_key: a region specific key for species codes and major benthic categories 

    * percent_cover_species: site level percent cover by species

    * percent_cover_site: site level percent cover by major benthic category

    * cover_strata: strata level mean percent cover of major benthic categories (unweighted)

    * Domain estimates: weighted regional (domain level) percent cover of major benthic categories

    * Domain_check: QAQC check that sums the regional percent cover. Should equal ~100
    
    * cover_check_site: QAQC check that sums the percent cover for each site. Should equal 100

Associated functions: NCRMP_make_weighted_LPI_data, load_NCRMP_benthic_cover_data
        
Example: 

```{r, echo=TRUE, warning=FALSE, message=FALSE, results='hide'}

require(ncrmp.benthics.analysis)

# Run function and assign names to dataframes 
tmp <- NCRMP_calculate_cover(region = "SEFCRI", project = "NCRMP")

# The function will store all of the function outputs as a list in tmp
# Convert list objects to dataframes and rename 
NCRMP_SEFCRI_2014_22_percent_cover_species <- tmp$percent_cover_species
NCRMP_SEFCRI_2014_22_percent_cover_site <- tmp$percent_cover_site
NCRMP_SEFCRI_2014_22_cover_group_key <- tmp$cover_group_key
NCRMP_SEFCRI_2014_22_cover_strata <- tmp$cover_strata
NCRMP_SEFCRI_2014_22_cover_region <- tmp$Domain_est
NCRMP_SEFCRI_2014_22_Domain_check <- tmp$Domain_check 
cover_check_site <- tmp$cover_check_site


```


### 1b. NCRMP_colony_percent_cover

__Calculates species-specific coral cover from at regional level (weighted) for individual years. Also produces a barplot of the most recent years data__


* parameters:
    * region: "SEFCRI", "FLK", "Tortugas", "STTSTJ", "STX", "PRICO", "GOM" 

    * ptitle: a string indicating the title of the plot
    
    * file_path: a string indicating the file path of stored barplot of most recent years' data

* outputs (dataframes):
    * region_means: weighted regional (domain level) percent cover of coral species
    
    * a barplot of weighted regional % cover by coral species for the most recent year's data, exported to the specified filepath

Associated functions: NCRMP_make_weighted_LPI_data, load_NCRMP_benthic_cover_data
        
Example: 

```{r, echo=TRUE, warning=FALSE, message=FALSE, results='hide'}

require(ncrmp.benthics.analysis)

# Run function and assign names to dataframes 
tmp <-NCRMP_colony_percent_cover(region = "Tortugas",
                                 ptitle = "The Dry Tortugas",
                                 file_path = "K:/_BioGeoProjects/NCRMP/Data Analysis/Analysis ready data/Figures" )


# The function will store all of the function outputs as a list in tmp
# Convert list objects to dataframes and rename 
NCRMP_Tort_2014_22_percent_cover_region_species <- tmp$region_means

```


### 2. NCRMP_make_weighted_LPI_data

__Applies weighting scheme to benthic cover data__ 

Analysis level strata varies between regions. In the Caribbean (USVI, PRICO), strata is represented by HABITAT CODE + DEPTH STRAT. In GOM, only one strata is used: FGBNMS, so essentially there is no weighting there. In FLorida, strata is a combination of STRAT + PROT. However, there is no PROT in SEFCRI. In FLK, PROT is always reset to 0 because the benthic sites are not allocated at the PROT level. In Tortugas, PROT is maintained as is. These are the recommended and default settings. 

This function is called within NCRMP_calculate_cover so the initial parameters you provide for region and year. will be passed on. 

Example of weighting application for Florida:

```{r}

#Calculate average cover (avcvr), stratum variance (svar), n and std at the strata and regional level using STRAT + PROT as the analysis strata. Inputdata = 1 stage site level cover for major benthic categories and ntot = the most recent NTOT file for the region, regardless of sampling year. 

# cover_est <- inputdata %>%
#    # group by analysis level strata
#   dplyr::mutate(ANALYSIS_STRATUM = paste(STRAT, "/ PROT =", PROT, sep = " ")) %>%
#   dplyr::group_by(ANALYSIS_STRATUM, STRAT, PROT, cover_group) %>% 
#   dplyr::summarise(
#     # calculate mean cover
#     avcvr = mean(Percent_Cvr),
#     # calculate stratum variance
#     svar = var(Percent_Cvr),
#     # calculate N
#     n = sum(n),
#     # calculate mean stratum depth
#     MIN_DEPTH = mean(MIN_DEPTH),
#     MAX_DEPTH = mean(MAX_DEPTH),
#     DEPTH_M = (MIN_DEPTH+MAX_DEPTH)/2) %>%
#   
#   # convert 0 for stratum variance so that the sqrt is a small # but not a 0
#   dplyr::mutate(svar = dplyr::case_when(svar == 0 ~ 0.00000001,
#                                         TRUE ~ svar)) %>%
#   dplyr::mutate(std = sqrt(svar))
# 
# cover_est <- cover_est %>%
#   # Merge ntot with cover_est
#   dplyr::full_join(., ntot) %>%
#   # stratum estimates
#   dplyr::mutate(whavcvr = wh * avcvr,
#                 whsvar = wh^2 * svar,
#                 whstd = wh * std,
#                 n = tidyr::replace_na(n, 0))
# 
# # Reformat data
# 
# # cover, unweighted by strata
# cover_strata <- cover_est %>%
#   dplyr::select(REGION, YEAR, ANALYSIS_STRATUM, STRAT, RUG_CD, PROT, DEPTH_M, NTOT, ngrtot, wh, cover_group, n, avcvr, svar, std) %>%
#   dplyr::mutate(METHOD = "UNWEIGHTED") %>%
#   dplyr::mutate(n = tidyr::replace_na(n, 0))
# 
# # Calculate Domain Estimates
#   Domain_est <- cover_est %>%
#     dplyr::group_by(REGION, YEAR, cover_group) %>%
#     dplyr::summarise(avCvr = sum(whavcvr),
#                      var = sum(whsvar, na.rm = T),
#                      std = sqrt(var),
#                      ngrtot = sum(NTOT) )



```

  Example of weighting application for Caribbean and GOM:

```{r}

# Calculate average cover (avcvr), stratum variance (svar), n and std at the strata and regional level using HABITAT + DEPTH_STRAT as the analysis strata. Inputdata = 1 stage site level cover for major benthic categories and ntot = the most recent NTOT file for the region, regardless of sampling year. 
# 
# cover_est <- inputdata %>%
#   # group by analysis level strata
#   dplyr::group_by(YEAR, ANALYSIS_STRATUM, STRAT, cover_group) %>% 
#   dplyr::summarise(
#     # compute average cover
#     avcvr = mean(Percent_Cvr),
#     # compute stratum variance
#     svar = var(Percent_Cvr),
#     # calculate N
#     n = sum(n),
#     # calculate mean stratum depth
#     MIN_DEPTH = mean(MIN_DEPTH),
#     MAX_DEPTH = mean(MAX_DEPTH),
#     DEPTH_M = (MIN_DEPTH+MAX_DEPTH)/2) %>%
#   # convert 0 for stratum variance so that the sqrt is a small # but not a 0
#   dplyr::mutate(svar = dplyr::case_when(svar == 0 ~ 0.00000001,
#                                         TRUE ~ svar)) %>%
#   dplyr::mutate(std = sqrt(svar))
# 
# cover_est <- cover_est %>%
#   # Merge ntot with cover_est
#   dplyr::full_join(., ntot) %>%
#   # stratum estimates
#   dplyr::mutate(whavcvr = wh * avcvr,
#                 whsvar = wh^2 * svar,
#                 whstd = wh * std,
#                 n = tidyr::replace_na(n, 0),
#                 # Add the following to match FL format
#                 PROT = NA,
#                 RUG_CD = NA)
# 
# # Reformat data
# 
# # cover, unweighted by strata
# cover_strata <- cover_est %>%
#   dplyr::select(REGION, YEAR, ANALYSIS_STRATUM, STRAT, RUG_CD, PROT, DEPTH_M, NTOT, ngrtot, wh, cover_group, n, avcvr, svar, std) %>%
#   dplyr::mutate(METHOD = "UNWEIGHTED") %>%
#   dplyr::mutate(n = tidyr::replace_na(n, 0))
# 
# 
# ## Calculate Domain Estimates
# Domain_est <- cover_est %>%
#   dplyr::group_by(REGION, YEAR, cover_group) %>%
#   dplyr::summarise(avCvr = sum(whavcvr),
#                    var = sum(whsvar, na.rm = T),
#                    std = sqrt(var),
#                    ngrtot = sum(NTOT) )

```




## Coral density

### 3a. NCRMP_DRM_calculate_colony_density

__Calculates colony density from site to regional level for all sampling years__


* parameters:
    * project: "NCRMP" (All regions), "NCRMP_DRM" (Florida regions only)

    * region: "SEFCRI", "FLK", "Tortugas", "STTSTJ", "STX", "PRICO", "GOM" 

    * species_filter: A string indicating whether a region-specific preset species filter should be applied to density calculations (species codes listed) or all species should be included (FALSE; the default).

* outputs (dataframes):

    * density_species: site level coral density by species
    
    * density_site: site level coral density

    * density_strata: strata level coral density (unweighted)

    * Domain estimates: weighted regional (domain level) coral density 

Associated functions: NCRMP_make_weighted_demo_data, load_NCRMP_DRM_demo_data

Example: 

```{r, echo=TRUE, warning=FALSE, message=FALSE, results='hide'}

require(ncrmp.benthics.analysis)

# Run function 
tmp <- NCRMP_DRM_calculate_colony_density(project = "NCRMP_DRM", 
                                           region = "FLK",
                                           species_filter = "FALSE")

# The function will store all of the function outputs as a list in tmp
# Convert list objects to dataframes and rename 

NCRMP_DRM_FLK_2014_22_density_species <- tmp$density_species
NCRMP_DRM_FLK_2014_22_density_site <- tmp$density_site
NCRMP_DRM_FLK_2014_22_density_strata <- tmp$density_strata
NCRMP_DRM_FLK_2014_22_density_region <- tmp$Domain_est

```


### 3b. NCRMP_DRM_colony_density_CV_and_occurrence

__Calculates colony density and coefficient of variation (CV) by species at the regional level for all sampling years. Also produces a barplot of occurrence and CV by species for the most recent years data__


* parameters:
    * project: "NCRMP" (All regions), "NCRMP_DRM" (Florida regions only)

    * region: "SEFCRI", "FLK", "Tortugas", "STTSTJ", "STX", "PRICO", "GOM" 

    * species_filter: A string indicating whether a region-specific preset species filter should be applied to density calculations (species codes listed) or all species should be included (FALSE; the default).
    
    * year: a numeric indicating the year of interest for plotting, usually the most recent year
    
    * ptitle: a string indicating the figure title
    
    * file_path: a string indicating the file path to store the produced figure

* outputs (dataframes):

    * region_means: weighted regional (domain level) coral density by species
    
    * region_means_cv20: weighted regional (domain level) coral density by species for species with a CV of 20 or less

Associated functions: NCRMP_make_weighted_demo_data, load_NCRMP_DRM_demo_data

Example: 

```{r, echo=TRUE, warning=FALSE, message=FALSE, results='hide'}

require(ncrmp.benthics.analysis)

# Run function 
tmp <- NCRMP_DRM_colony_density_CV_and_occurrence(region = "Tortugas",
                                              ptitle = "Dry Tortugas",
                                              year = 2022,
                                              file_path = "K:/_BioGeoProjects/NCRMP/Data Analysis/Analysis ready data/Figures" ,  #must be connected to VPN if filepath is to a network drive
                                              species_filter = "NULL",
                                              project = "NCRMP")


# The function will store all of the function outputs as a list in tmp
# Convert list objects to dataframes and rename 

NCRMP_Tort_Occ_Den_CV <- tmp$region_means
NCRMP_Tort_Occ_Den_CV_20 <- tmp$region_means_cv20

```


### 4. NCRMP_make_weighted_demo_data

__Applies weighting scheme to coral demographic data (coral density, species richness, mortality, disease/bleaching prevalence).__ 

Analysis level strata varies between regions. In the Caribbean (USVI, PRICO), strata is represented by HABITAT CODE + DEPTH STRAT. In GOM, only one strata is used: FGBNMS, so essentially there is no weighting there. In FLorida, strata is a combination of STRAT + PROT. However, there is no PROT in SEFCRI. In FLK, PROT is always reset to 0 because the benthic sites are not allocated at the PROT level. In Tortugas, PROT is maintained as is. These are the recommended and default settings. 

This function is called within NCRMP_DRM_calculate_colony_density so the initial parameters you provide for region and year will be passed on. 

Example of weighting application for Florida:

```{r}

# Calculate average density (avden), stratum variance (svar), n and std at the strata and regional level using STRAT + PROT as the analysis strata. Inputdata = 1 stage site level coral density and ntot = the most recent NTOT file for the region, regardless of sampling year. 

# density_est <- inputdata %>%
#   # group by analysis level strata
#   dplyr::mutate(ANALYSIS_STRATUM = paste(STRAT, "/ PROT =", PROT, sep = " ")) %>%
#   dplyr::group_by(YEAR, ANALYSIS_STRATUM, STRAT, PROT) %>% 
#   dplyr::summarise(
#     # compute average density
#     avden = mean(DENSITY),
#     # compute stratum variance
#     svar = var(DENSITY),
#     # calculate N
#     n = length(DENSITY)) %>%
#   # convert 0 for stratum variance so that the sqrt is a small # but not a 0
#   dplyr::mutate(svar = dplyr::case_when(svar == 0 ~ 0.00000001,
#                                         TRUE ~ svar)) %>%
#   dplyr::mutate(std = sqrt(svar))
# 
# density_est <- density_est %>%
#   # Merge ntot with coral_est_spp
#   dplyr::full_join(., ntot) %>%
#   # stratum estimates
#   dplyr::mutate(whavden = wh * avden,
#                 whsvar = wh^2 * svar,
#                 whstd = wh * std,
#                 n = tidyr::replace_na(n, 0))
# 
# # Reformat output
# 
# density_strata <-  density_est %>%
#   dplyr::select(REGION, YEAR, ANALYSIS_STRATUM, STRAT, RUG_CD, PROT, NTOT, ngrtot, wh, n, avden, svar, std)
# 
# # Calculate Domain Estimates
# Domain_est <- density_est %>%
#   dplyr::group_by(REGION, YEAR) %>%
#   dplyr::summarise(avDen = sum(whavden),
#                    var = sum(whsvar, na.rm = T),
#                    std = sqrt(var),
#                    ngrtot = sum(NTOT) )



```


Example of weighting application for Caribbean:

```{r}

# Calculate average density (avden), stratum variance (svar), n and std at the strata and regional level using HABITAT + DEPTH_STRAT as the analysis strata. Inputdata = 1 stage site level coral density and ntot = the most recent NTOT file for the region, regardless of sampling year. 
# 
# # Calculate avdns, svar, n and std
# density_est <- inputdata %>%
#   # group by analysis level strata
#   dplyr::mutate(ANALYSIS_STRATUM = STRAT) %>%
#   dplyr::group_by(YEAR, ANALYSIS_STRATUM, STRAT) %>% 
#   dplyr::summarise(# compute average density
#     avden = mean(DENSITY),
#     # compute stratum variance
#     svar = var(DENSITY),
#     # calculate N
#     n = length(DENSITY)) %>%
#   # convert 0 for stratum variance so that the sqrt is a small # but not a 0
#   dplyr::mutate(svar = dplyr::case_when(svar == 0 ~ 0.00000001,
#                                         TRUE ~ svar)) %>%
#   dplyr::mutate(std = sqrt(svar))
# 
# density_est <- density_est %>%
#   # Merge ntot with coral_est_spp
#   dplyr::full_join(., ntot) %>%
#   # stratum estimates
#   dplyr::mutate(whavden = wh * avden,
#                 whsvar = wh^2 * svar,
#                 whstd = wh * std,
#                 n = tidyr::replace_na(n, 0),
#                 # Add the following to match FL format 
#                 PROT = NA,           
#                 RUG_CD = NA)
# 
# # Reformat output
# 
# density_strata <-  density_est %>%
#   dplyr::select(REGION, YEAR, ANALYSIS_STRATUM, STRAT, RUG_CD, PROT, NTOT, ngrtot, wh, n, avden, svar, std)
# 
# # Calculate Domain Estimates
# Domain_est <- density_est %>%
#   dplyr::group_by(REGION, YEAR) %>%
#   dplyr::summarise(avDen = sum(whavden),
#                    var = sum(whsvar, na.rm = T),
#                    std = sqrt(var),
#                    ngrtot = sum(NTOT) )


```

       


## Species richness and diversity

### 5. NCRMP_DRM_calculate_species_richness_diversity

__Calculates species richness from site to regional level for all sampling years. Calculates Simpson, Inverse Simpson and Shannon–Wiener diversity indices from site to regional level for all sampling years using the R package vegan.__


* parameters:
    * project: "NCRMP" (All regions), "NCRMP_DRM" (Florida regions only)

    * region: "SEFCRI", "FLK", "Tortugas", "STTSTJ", "STX", "PRICO", "GOM" 

* outputs (dataframes):

    * species_list: a list of a species present for all sampling years 

    * richness_site: site level species richness. Includes juveniles. Note there is no accouting for transect length/effort

    * richness_strata: strata level mean species richness (unweighted). Includes juveniles. Note there is no accouting for transect length/effort

    * Domain estimates: weighted regional (domain level) species richness. Includes juveniles. Note there is no accouting for transect length/effort

    * species_diversity_site: site level species diversity using the Simpson, Inverse Simpson and Shannon-Weiner diversity indices. Based only on adults since the abundance of juveniles is not recorded. Note there is no accouting for transect length/effort

    * species_diversity_strata: strata level species diversity using the Simpson, Inverse Simpson and Shannon-Weiner diversity indices (unweighted). Based only on adults since the abundance of juveniles is not recorded. Note there is no accouting for transect length/effort
    
    * species_diversity_region: weighted regional (domain level) species diversity using the Simpson, Inverse Simpson and Shannon-Weiner diversity indices. Based only on adults since the abundance of juveniles is not recorded. Note there is no accouting for transect length/effort
    
Associated functions: NCRMP_make_weighted_demo_data

Example: 

```{r, echo=TRUE, warning=FALSE, message=FALSE, results='hide'}

require(ncrmp.benthics.analysis)

# Run function 
tmp <- NCRMP_DRM_calculate_species_richness_diversity(project = "NCRMP",
                                                       region= "Tortugas")

# The function will store all of the function outputs as a list in tmp
# Convert list objects to dataframes and rename 

NCRMP_Tort_2014_22_spp_list <- tmp$species_list 
NCRMP_Tort_2014_22_richness_site <- tmp$richness_site
NCRMP_Tort_2014_22_richness_strata <- tmp$richness_strata
NCRMP_Tort_2014_22_richness_region <- tmp$Domain_est

NCRMP_Tort_2014_22_diversity_site <- tmp$species_diversity_site
NCRMP_Tort_2014_22_diversity_strata <- tmp$diversity_strata
NCRMP_Tort_2014_22_diversity_region <- tmp$Domain_est_div

```

Regional species richness and diversity are calculated using the same weighting scheme as coral density. See NCRMP_make_weighted_demo_data and NCRMP_make_weighted_demo_data_RC for example code. Only the default analysis strat settings are available for species diversity. See https://cran.r-project.org/web/packages/vegan/vegan.pdf for more information on the diversity indices. 


## Mortality

### 6. NCRMP_DRM_calculate_mortality

__Calculates old and recent mortality (percent of colony surface) from species to regional level for all sampling years__

* parameters:
    * project: "NCRMP" (All regions), "NCRMP_DRM" (Florida regions only)

    * region: "SEFCRI", "FLK", "Tortugas", "STTSTJ", "STX", "PRICO", "GOM" 

    * species_filter: A string indicating whether a region-specific preset species filter should be applied to density calculations (species codes in a list) or all species should be included (FALSE; the default). Generally not used as species-specific estimates are also provided by default.

* outputs (dataframes):

    * old_mortality_site: site level old mortality (across all corals)
    
    * recent_mortality_site: site level recent mortality (across all corals)

    * old_mortality_strata: strata level mean old mortality (unweighted, across all corals)
    
    * rec_mortality_strata: strata level mean recent mortality (unweighted, across all corals)
    
    * old_mortality_species_strata: strata level mean old mortality (unweighted, by species)
    
    * rec_mortality_species_strata: strata level mean recent mortality (unweighted, by species)

    * Domain_est_old_mort: weighted regional (domain level) old mortality (across all species)
    
    * Domain_est_rec_mort: weighted regional (domain level) recent mortality (across all species)
    
    * Domain_est_old_mort_species: weighted regional (domain level) old mortality, by species. Not species-specific mortality weighting is based on strata where species is PRESENT
    
    * Domain_est_rec_mort_species: weighted regional (domain level) recent mortality, by species. Not species-specific mortality weighting is based on strata where species is PRESENT
    
    * ntot_check_rec_mort_species: QAQC check to ensure species-specific NTOT's add up to 1
    
    * ntot_check_old_mort_species: QAQC check to ensure species-specific NTOT's add up to 1

Associated functions: NCRMP_make_weighted_demo_data

Example:

```{r, echo=TRUE, warning=FALSE, message=FALSE, results='hide'}

require(ncrmp.benthics.analysis)

# Run function
tmp <- NCRMP_DRM_calculate_mortality(project = "NCRMP",
                                      region = "GOM",
                                      species_filter = "FALSE")

# The function will store all of the function outputs as a list in tmp
# Convert list objects to dataframes and rename

NCRMP_FGBNMS_2013_22_old_mort_site <- tmp$old_mortality_site
NCRMP_FGBNMS_2013_22_rec_mort_site <- tmp$recent_mortality_site

NCRMP_FGBNMS_2013_22_old_mort_strata <- tmp$old_mortality_strata
NCRMP_FGBNMS_2013_22_rec_mort_strata <- tmp$rec_mortality_strata

NCRMP_FGBNMS_2013_22_old_mort_region <- tmp$Domain_est_old_mort
NCRMP_FGBNMS_2013_22_rec_mort_region <- tmp$Domain_est_rec_mort

NCRMP_FGBNMS_2013_22_old_mort_species_region <- tmp$Domain_est_old_mort_species
NCRMP_FGBNMS_2013_22_rec_mort_species_region <- tmp$Domain_est_rec_mort_species
# check ntot weighting for species-specific mortality - should only be values of 1 (sum of weights across strata for each species in each year)
unique(tmp$ntot_check_rec_mort_species$wh_sum)
unique(tmp$ntot_check_old_mort_species$wh_sum)

```

Regional old and recent mortality are calculated using the same weighting scheme as coral density. See NCRMP_make_weighted_demo_data for example code. Species-specific mortality estimates are done based on strata a species is PRESENT. See load_NTOT_species for code. 



## Invertebrate density 

### 7. NCRMP_calculate_invert_density

__Calculates *Diadema antillarum* density from site to regional level for all sampling years__

* parameters:

    * region: "SEFCRI", "FLK", "Tortugas", "STTSTJ", "STX", "PRICO", "GOM" 
    
    * project: "NCRMP" (default) or "MIR" (in development)

* outputs (dataframes):

    * diadema_density_site: site level Diadema density

    * invert_strata: strata level mean Diadema density (unweighted)

    * Domain_est: weighted regional (domain level) Diadema density


Associated functions: NCRMP_make_weighted_invert_density

Example: 

```{r, echo=TRUE, warning=FALSE, message=FALSE, results='hide'}

require(ncrmp.benthics.analysis)


# Run function 
tmp <- NCRMP_calculate_invert_density(region = "PRICO", project = "NCRMP")

# The function will store all of the function outputs as a list in tmp
# Convert list objects to dataframes and rename 

NCRMP_PRICO_2014_21_invert_den_site <- tmp$diadema_density_site
NCRMP_PRICO_2014_21_invert_den_strata <- tmp$invert_strata # Right now just Diadema
NCRMP_PRICO_2014_21_invert_den_region <- tmp$Domain_est



```

Regional *Diadema* density is calculated using the same weighting scheme as coral density. See NCRMP_make_weighted_invert_density for example code. This function is currently only calculating *Diadema* density. Future updates will include conch and lobster; but until these updates are complete, please see the code in this function for guidance on how to calculate density for the other invertebrates.  

## Disease/bleaching prevalence

### 8a. NCRMP_DRM_calculate_disease_prevalence_colonies

__Calculates disease and bleaching prevalence at the site (% colonies per site), site and species (% colonies per site per species), strata (mean of % colonies per site), and regional (weighted mean of % colonies per site) level for all sampling years__

* parameters:
    * project: "NCRMP_DRM" (Florida regions only), "NCRMP" (All regions)

    * region: "SEFCRI", "FLK", "Tortugas", "STTSTJ", "STX", "PRICO", "GOM" 

    * species_filter: A string indicating whether a region-specific preset species filter should be applied to density calculations (species codes in a list) or all species should be included (FALSE; the default). Note the species filter is not used as much as species-specific domain estimates are also provided.

* outputs (dataframes):

    * dis_ble_prev_species: site and species level disease and bleaching prevalence
    
    * dis_ble_prev_site: site level disease and bleaching prevalence 

    * dis_prev_strata: strata level disease prevalence (unweighted)
    
    * ble_prev_strata: strata level bleaching prevalence (unweighted)

    * Domain_est: weighted regional (domain level) disease and bleaching prevalence

Associated functions: NCRMP_make_weighted_demo_data

Example:

```{r, echo=TRUE, warning=FALSE, message=FALSE, results='hide'}

require(ncrmp.benthics.analysis)

# Run function 

tmp <- NCRMP_DRM_calculate_disease_prevalence_colonies(project = "NCRMP_DRM",
                                                       region = "SEFCRI",
                                                       species_filter = "FALSE")

# The function will store all of the function outputs as a list in tmp
# Convert list objects to dataframes and rename 

NCRMP_DRM_SEFCRI_2014_22_dis_ble_prev_species <- tmp$dis_ble_prev_species
NCRMP_DRM_SEFCRI_2014_22_dis_ble_prev_site <- tmp$dis_ble_prev_site
NCRMP_DRM_SEFCRI_2014_22_dis_prev_strata <- tmp$dis_prev_strata
NCRMP_DRM_SEFCRI_2014_22_ble_prev_strata <- tmp$ble_prev_strata
NCRMP_DRM_SEFCRI_2014_22_dis_ble_prev_region <- tmp$Domain_est

```

Regional disease prevalence are calculated using the same weighting scheme as coral density. See NCRMP_make_weighted_demo_data for example code. 

Note: Disease was not sampled in the Flower Garden Banks National Marine Sanctuary (FGBNMS) in 2013


### 8b. NCRMP_DRM_calculate_dis_ble_prevalence_species_domain

__Calculates disease and bleaching prevalence by species at the regional (weighted mean of % colonies per site) level for all sampling years. Note weighting is specific to the strata each species is present in, because disease prevalence is NA (not 0) in a strata where a species is absent__

* parameters:
    * project: "NCRMP_DRM" (Florida regions only), "NCRMP" (All regions)

    * region: "SEFCRI", "FLK", "Tortugas", "STTSTJ", "STX", "PRICO", "GOM" 

    * species_filter: A string indicating whether a region-specific preset species filter should be applied to density calculations (species codes in a list) or all species should be included (FALSE; the default). Note the species filter is not used as much as species-specific domain estimates are also provided.

* outputs (dataframes):

    * DomainEst_dis: weighted regional (domain level) disease prevalence by species
    
    * DomainEst_bl: weighted regional (domain level) bleaching prevalence by species

    * wh_total: QAQC check to ensure species-specific NTOT's add up to 1

Associated functions: NCRMP_make_weighted_demo_data

Example:

```{r, echo=TRUE, warning=FALSE, message=FALSE, results='hide'}

require(ncrmp.benthics.analysis)

# Run function 

tmp <- NCRMP_DRM_calculate_dis_ble_prevalence_species_domain(project="NCRMP",
                                                             region = "STX")
# ntot check - wh_total should always be 1 (sum of weights across strata for each species in each year)
unique(tmp$ntot_check$wh_total)

# The function will store all of the function outputs as a list in tmp
# Convert list objects to dataframes and rename 

NCRMP_STX_2015_21_dis_prev_species_region <- tmp$DomainEst_dis
NCRMP_STX_2015_21_ble_prev_species_region <- tmp$DomainEst_bl

```


### 8c. NCRMP_DRM_calculate_disease_prevalence_sites

__Calculates number of sites with disease and bleaching at the strata and regional level for all sampling years__

* parameters:
    * project: "NCRMP_DRM" (Florida regions only), "NCRMP" (All regions)

    * region: "SEFCRI", "FLK", "Tortugas", "STTSTJ", "STX", "PRICO", "GOM" 

    * species_filter: A string indicating whether a region-specific preset species filter should be applied to density calculations (species codes in a list) or all species should be included (FALSE; the default). Note the species filter is not used as much as species-specific domain estimates are also provided.

* outputs (dataframes):

    * dis_ble_prev_site: site level disease and bleaching presence
    
    * dis_ble_prev_strata: number of sites in each strata with disease and bleaching

    * disease_prev_region: number of sites across the region with disease and bleaching

Associated functions: NCRMP_make_weighted_demo_data

Example:

```{r, echo=TRUE, warning=FALSE, message=FALSE, results='hide'}

require(ncrmp.benthics.analysis)

# Run function 

tmp <- NCRMP_DRM_calculate_disease_prevalence_sites(project = "NCRMP",
                                                    region = "STTSTJ",
                                                    species_filter = "FALSE")

# The function will store all of the function outputs as a list in tmp
# Convert list objects to dataframes and rename 

NCRMP_STTSTJ_2013_21_dis_ble_PA_site <- tmp$dis_ble_prev_site
NCRMP_STTSTJ_2013_21_dis_n_sites_per_strata <- tmp$dis_ble_prev_strata
NCRMP_STTSTJ_2013_21_dis_ble_n_sites_per_region <- tmp$disease_prev_region

```




# Functions for all regions and years

### These functions create outputs that contain all regions and all years. They are not weighted. 

## ESA listed corals - presence absence

### 9. NCRMP_calculate_ESA_corals_PresAbs

__Calculates strata and site level presence / absence for all regions and sampling years. Currently NCRMP data only.__

* parameters:
    * None

* outputs (dataframes):

    * NCRMP_AllRegions_Years_ESA_PresAbs_Strat: strata level ESA presence/absence
    
    * NCRMP_AllRegions_Years_ESA_PresAbs_Site: site level ESA presence/absence

    * region_year: a list to check that every region and year is included

Associated functions: None

Example:
```{r}
require(ncrmp.benthics.analysis)

# Run function 
tmp <- NCRMP_calculate_ESA_corals_PresAbs()

# The function will store all of the function outputs as a list in tmp
# Convert list objects to dataframes and rename 

NCRMP_AllRegions_Years_ESA_PresAbs_Strat <- tmp$NCRMP_AllRegions_Years_ESA_PresAbs_Strat
NCRMP_AllRegions_Years_ESA_PresAbs_Site <- tmp$NCRMP_AllRegions_Years_ESA_PresAbs_Site

region_year <- tmp$region_year

```



### 10. NCRMP_make_ESA_corals_FOI

__Calculates regional ESA frequency of incidence, across all ESA species and for each, for all regions and sampling years based on the number of sites ESA are present. Currently NCRMP data only.__

* parameters:
    * None

* outputs (dataframes):

    * FOI_region: regional frequency of incidence across all ESA corals for all regions and years
    
    * FOI_species: regional frequency of incidence of ESA corals by species for all regions and years

Associated functions: None

Example:
```{r}
require(ncrmp.benthics.analysis)

# Run function 
tmp <- NCRMP_make_ESA_corals_FOI()

# The function will store all of the function outputs as a list in tmp
# Convert list objects to dataframes and rename 
NCRMP_ESA_FOI_region <- tmp$FOI_region
NCRMP_ESA_FOI_species <- tmp$FOI_species

```



## Bleaching abundance

### 11. NCRMP_make_bleaching_abundance

__Calculates species level bleaching abundance by each transect for all regions and sampling years. Currently NCRMP data only__


* parameters:
    * None

* outputs (dataframes):

    * NCRMP_bleaching_abundance_Allyears_Allregions: species level bleaching abundance for each transect

Associated functions: None

Example:

```{r}
require(ncrmp.benthics.analysis)

# Run function 

tmp <- NCRMP_make_bleaching_abundance()

# The function will store all of the function outputs as a list in tmp
# Convert list objects to dataframes and rename 

NCRMP_bleaching_abundance_Allyears_Allregions <- tmp$NCRMP_bleaching_abundance_Allyears_Allregions

```



______________________________________________________________________________________________________



# Frequently asked questions:

### 1. What species are used when you filter the species in NCRMP_DRM_calculate_colony_density, NCRMP_DRM_calculate_disease_prevalence or NCRMP_DRM_calculate_mortality?

# NCRMP species filter developed in status report meetings

   Report_card_filter <- c("ACR CERV", "ACR PALM", "COL NATA", "DIP LABY", "MON CAVE", "ORB ANNU", "ORB FRAN", "ORB FAVE",
                           "POR PORI", "PSE STRI", "PSE CLIV", "SID SIDE", "STE INTE")
                           
### 2. Which benthic categories are combined into "OTHER" in NCRMP_calculate_cover?

The following major benthic categories are currently classified as OTHER in the benthic cover metrics: SUBSTRATE, HYDROCORALS, OTHER INVERTEBRATES, CYANOBACTERIA, and SEAGRASS. Please see ncrmp_frrp_sppcodes, which is contained in the package, for information on how taxa are categorized into major benthic categories.  

# Disclaimer

This repository is a scientific product and is not official communication of the National Oceanic and
Atmospheric Administration, or the United States Department of Commerce. All NOAA GitHub project
code is provided on an ‘as is’ basis and the user assumes responsibility for its use. Any claims against the Department of Commerce or Department of Commerce bureaus stemming from the use of this GitHub project will be governed by all applicable Federal law. Any reference to specific commercial products, processes, or services by service mark, trademark, manufacturer, or otherwise, does not constitute or imply their endorsement, recommendation or favoring by the Department of Commerce. The Department of Commerce seal and logo, or the seal and logo of a DOC bureau, shall not be used in any manner to imply endorsement of any commercial product or activity by DOC or the United States Government.

